---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ABDtools

<!-- badges: start -->
<!-- badges: end -->

This is a companion package for the Africa Bird Data packages [`ABAP`](https://github.com/AfricaBirdData/ABAP)
and [`CWAC`](https://github.com/AfricaBirdData/CWAC). These packages allow us to 
pull African bird citizen science data from the projects of the same name to our
R session. For now, the `ABDtools` package adds the functionality
necessary to annotate these data with environmental information from the 
[Google Earth Engine data catalog](https://developers.google.com/earth-engine/datasets).
This should make the analysis of ABAP and CWAC data easier and more reproducible.

However, there is nothing preventing you from using the `ABDtools` package to 
annotate other types of data. As long as the data comes in a spatial format,
as either points or polygons, we can use `ABDtools` to annotate them with data
from the GEE catalog.


## HOW TO ANNOTATE DATA WITH ABDtools

In the instructions below we assume that you have a basic understanding of how
the [`ABAP`](https://github.com/AfricaBirdData/ABAP) and 
[`CWAC`](https://github.com/AfricaBirdData/CWAC) packages work and the type of
data they provide. If you don't, please visit their repository pages, and read the
use cases.

The only difference between ABAP and CWAC data with respect to how ABDtools works
is that each ABAP data point refers to a polygon (pentad) and each CWAC data point
refers to a point in space. All GEE data comes in the form of images (pixels),
so each spatial point from CWAC corresponds to a single pixel in a GEE image, but
an ABAP pentad contains multiple image pixels. Therefore, we need a way to summarise
the values of the different pixels contained in a polygon. We will use functions,
which in GEE jargon are called "reducers", such as "mean" or "count". Keep this
in mind when reading through the next sections.


### Installation

Install the latest version of `ABDtools`, which for now is only present on GitHub.

```{r, eval=FALSE}
install.packages("remotes") # In case this is not already installed
remotes::install_github("AfricaBirdData/ABDtools")
```

The package `ABDtools` builds upon [`rgee`](https://github.com/r-spatial/rgee),
which translates R code into Python code using `reticulate`, and allows us
to use the Google Earth Engine (GEE) Python client libraries from R! You can find extensive documentation
about `rgee` [here](https://github.com/r-spatial/rgee).

But first, we need to create a GEE account, install `rgee` (and dependencies)
and configure our Google Drive to upload and download data to and from GEE.
There are other ways of upload and download but we recommend Google Drive,
because it is free, simple and effective. Configuration is a bit of a process, but you
will have to do this only once. 

 - To create a GEE account, please follow these [instructions](https://earthengine.google.com/signup/).
 - To install `rgee`, follow these [instructions](https://github.com/r-spatial/rgee#installation).
 - To configure Google Drive, follow these [instructions](https://r-spatial.github.io/rgee/articles/rgee01.html).
 
We have nothing to do with the above steps, so if you get stuck along the way,
please search the web or contact the developers of these packages directly.

Well done if you managed that! With configuration out of the way, let's
see how to annotate some data. We've coded some wrappers around basic
functions from the `rgee` package to provide some **basic** functionality without
having to know almost anything about GEE. However, if you want more complicated
workflows, we totally recommend learning how to use GEE and `rgee` and exploit 
their enormous power.


### Initialize

Most image processing and handling of spatial features happens in GEE servers.
Therefore, there will be constant flow of information between our computer
(client) and GEE servers (server). This information flow happens through
Google Drive. So when we start our session we need to initialize a connection
with GEE and Google Drive.

```{r eval=TRUE}
# Initialize Earth Engine
library(rgee)

# Check installation
ee_check()

# Initialize rgee and Google Drive
ee_Initialize(drive = TRUE)
```

Make sure that all tests and checks are passed. If so, you are good to go!


### Uploading data to GEE

Firstly, you will need to upload the data you want to annotate to GEE. These
data will go to your 'assets' directory in the GEE server and it will stay there
until you remove it. So if you have uploaded some data, you don't have to 
upload it again. 

GEE-related functions from the `ADBtools` package work with spatial data and therefore
our detection data must be uploaded as spatial objects.

For example, to upload all ABAP pentads in the North West province of South Africa
(these are already on an `sf` format out of the box!), we can use:

```{r eval=FALSE}
library(ABAP)
library(sf)
library(dplyr, warn.conflicts = FALSE)
library(ABDtools)

# Load ABAP pentads
pentads <- getRegionPentads(.region_type = "province", .region = "North West")

# Set an ID for your remote asset (data in GEE)
assetId <- file.path(ee_get_assethome(), 'pentads')

# Upload to GEE (if not done already - do this only once)
uploadFeaturesToEE(feats = pentads,
                   asset_id = assetId,
                   load = FALSE)

# Load the remote asset to you local computer to work with it
ee_pentads <- ee$FeatureCollection(assetId)

```

Now, the object `pentads` lives in your machine, but the object `ee_pentads` lives
in the GEE server. You only have a "handle" to it in your machine to manipulate it.
This might seem a bit confusing at first but you will get used to it.

We are now going to also upload some CWAC count data to see how annotating these
data compares to ABAP data.

GEE can be temperamental with the type of data you upload, so we recommend leaving
in the data frame only those variables, that you will be needing for your GEE analysis.
Remember to always include an identifier field that allows you to join the results
back to you original data. In this case, we will leave only the start
date of the survey and the aforementioned identifier. GEE likes dates in a character
format, so lets transform the variable before proceeding.

```{r eval=FALSE}

# Select variables and format them
counts_to_upload <- bd_counts %>%
  dplyr::select(ID, StartDate) %>%
  mutate(StartDate = as.character(StartDate))

# Set an ID for your remote asset (data in GEE)
assetId <- file.path(ee_get_assethome(), 'my_cwac_counts')

# Upload to GEE (if not done already - do this only once)
uploadFeaturesToEE(feats = counts_to_upload,
                   asset_id = assetId,
                   load = FALSE)

# Load the remote asset to you local computer to work with it
ee_counts <- ee$FeatureCollection(assetId)

```


### Annotate data with a GEE image

An image in GEE jargon is the same thing as a raster in R. There are also image
collections which are like raster stacks (we'll see more about these later). You 
can find a full catalog of what is available in GEE [here](https://developers.google.com/earth-engine/datasets/catalog).
If you want to use data from a single image you can use the function
`addVarEEimage()`.

For example, let's annotate our ABAP pentads with 
[surface water occurrence](https://developers.google.com/earth-engine/datasets/catalog/JRC_GSW1_3_GlobalSurfaceWater),
which is the frequency with which water was present in each pixel. We'll need
the name of the layer in GEE, which is given in the field "Earth Engine Snippet".
Images can have multiple bands -- in this case, we select "occurrence".

Finally, images have pixels but our spatial objects are polygons, so we need to
define some type of summarizing function. The same thing that `raster::extract()`
would need. In GEE this is called a "reducer". In this case, we will select the
"mean" function (i.e., mean water occurrence per pixel within each pentad).


```{r eval=FALSE}

pentads_water <- addVarEEimage(ee_feats = ee_pentads,                   # Note that we need our remote asset here
                               image = "JRC/GSW1_3/GlobalSurfaceWater",   # You can find this in the code snippet
                               reducer = "mean",
                               bands = "occurrence")

```


If we were to annotate CWAC counts, everything would be a bit easier, because
counts are not associated with polygons but with specific locations in space
(i.e., points).
 
```{r eval=FALSE}
counts_water <- addVarEEimage(ee_feats = ee_counts,                   # Note that we need our remote asset here
                              image = "JRC/GSW1_4/GlobalSurfaceWater",   # You can find this in the code snippet
                              bands = "occurrence")

```

Note that now we didn't need to include a `reducer` argument, because we don't 
need to summarise multiple pixels enclosed by a polygon.


### Annotate data with a GEE collection

Sometimes environmental data don't come in a single image, but in multiple images.
For example, we might have one image for each day, week, month, etc. Again, we can check
all available data in the [GEE catalog](https://developers.google.com/earth-engine/datasets/catalog).
When we want to annotate data with a collection, we have two options:

 - We can use `addVarEEcollection()` to summarize the image collection over time
 (e.g., calculate the mean over a period of time)
 
 - Or we can use `addVarEEclosestImage()` to annotate each row in our bird data with
 the image in the collection that is closest in time. This is particularly useful
 to annotate visit data, where we are interested in the conditions observers found
 during their surveys. 
 
We demonstrate both options above by annotating data with the
[TerraClimate dataset](https://developers.google.com/earth-engine/datasets/catalog/IDAHO_EPSCOR_TERRACLIMATE),
which provides monthly climate data. If we were interested in annotating pentad
data with the mean minimum temperature across the year 2010, we could use

```{r eval=FALSE}

pentads_tmmn <- addVarEEcollection(ee_feats = ee_pentads,                    # Note that we need our remote asset here
                                   collection = "IDAHO_EPSCOR/TERRACLIMATE",   # You can find this in the code snippet
                                   dates = c("2010-01-01", "2011-01-01"),
                                   temp_reducer = "mean",
                                   spt_reducer = "mean",
                                   bands = "tmmn")

```

Note that in this case, we had to specify a temporal reducer `temp_reducer`to summarize
pixel values over time. A reducer in GEE is a function that summarizes temporal or spatial data.
The `dates` argument subsets the whole TerraClimate dataset to those
images between `dates[1]` (inclusive) and `dates[2]` (exclusive). Effectively, the
function computes a summary of the values of each pixel (a mean, in this case) across
all images (i.e., months), to create a single image. Then, it uses this new image
to annotate our data.

We then, use the argument `spt_reducer` to specify a function to summarise pixel
values contained in each pentad (a spatial summary, as opposed to the previous 
temporal summary). As explained earlier, this spatial summary is not necessary
when working with point data, such as CWAC counts.

If we wanted to annotate data with the closest image in the collection, 
instead of with a temporal summary, then we would need to upload to GEE data with an
associated date. Dates must be in a character format ("yyyy-mm-dd") and the
variable must be called 'Date' (case sensitive). We already did some of this at
the beginning (please check the section 'Uploading data to GEE' if you can't remember),
but we will need to adjust our data slightly to work with TerraClimate data.

TerraClimate offers monthly data and the date associated with each image is always the
first day of the month. This means that if we have data corresponding to a date after
the 15th of the month they will be matched against the next month, because 
that's the one closest in time.

Each image collection has its own convention, and we must check what is appropriate in each
case. Here, for illustration purposes, we will change all dates in our data to be on the first
of the month to match TerraClimate.

As an example, let's download ABAP data for the Maccoa Duck in 2010 and
annotate these data with TerraClimate's minimum temperature data.

```{r eval=FALSE}

# Load ABAP pentads
pentads <- getRegionPentads(.region_type = "country", .region = "South Africa")

# Download Maccoa Duck
id <- searchAbapSpecies("Duck") %>% 
    filter(Common_species == "Maccoa") %>% 
    pull(SAFRING_No)

visit <- getAbapData(.spp_code = id,
                     .region_type = "country",
                     .region = "South Africa",
                     .years = 2008)

# Make spatial object
visit <- visit %>% 
  left_join(pentads, by = c("Pentad" = "Name")) %>% 
  st_sf() %>% 
  filter(!st_is_empty(.))   # Remove rows without geometry

# NOTE: TerraClimate offers monthly data. The date of each image is the beginning of
# the month, which means that dates after the 15th will be matched against the
# next month. I will change all dates to be on the first of the month for the
# analysis
visit <- visit %>% 
    dplyr::select(CardNo, StartDate, Pentad, TotalHours, Spp) %>% 
    mutate(Date = lubridate::floor_date(StartDate, "month"))

# Load to EE (if not done already)
assetId <- file.path(ee_get_assethome(), 'visit2008')

# Format date and upload to GEE
visit %>%
    dplyr::select(CardNo, Pentad, Date) %>%
    mutate(Date = as.character(Date)) %>%   # GEE doesn't like dates
    sf_as_ee(assetId = assetId,
             via = "getInfo_to_asset")

# Load the remote data asset
ee_visit <- ee$FeatureCollection(assetId)

# Annotate with GEE TerraClimate
visit_new <- addVarEEclosestImage(ee_feats = ee_visit,
                                  collection = "IDAHO_EPSCOR/TERRACLIMATE",
                                  reducer = "mean",                          # We only need spatial reducer
                                  maxdiff = 15,                              # This is the maximum time difference that GEE checks
                                  bands = c("tmmn"))

```


### Convert an image collection to a multi-band image

Lastly, we have made a convenience function that converts an image collection
into a multi-band image. This is useful because you can only annotate one
image at a time, but all the bands in the image get annotated. So if you want to
add several variables to your data, you can first create a multi-band image and 
then annotate with all bands at once. In this way you minimize the traffic between your
machine and GEE servers saving precious time and bandwidth.

Here we show how to find the mean NDVI for each year between 2008 and 2010,
create a multi-band image and annotate our data with these bands.

```{r eval=FALSE}

# Create a multi-band image with mean NDVI for each year
multiband <- EEcollectionToMultiband(collection = "MODIS/006/MOD13A2",
                                     dates = c("2008-01-01", "2020-01-01"),
                                     band = "NDVI",                       # You can find what bands are available from GEE catalog
                                     group_type = "year",
                                     groups = 2008:2019,
                                     reducer = "mean",
                                     unmask = FALSE)

# Find mean (mean) NDVI for each pentad and year
pentads_ndvi <- addVarEEimage(ee_feats = ee_pentads,
                              image = multiband,
                              reducer = "mean")

```

CWAC count data would be annotated in the exact same way but without any
`reducer` (`reducer = NULL`).

### Use reference polygons to annotate CWAC data

Up until now, we have been using the centroid of the CWAC site as a reference to
annotate our counts with. However, we might be interested in taking a broader
reference area. For example, we might want to use the boundaries of the CWAC site
where the counts were collected. We can use any other polygon we want, like the 
catchment the wetland belongs to, for example. Which polygon
to use will depend on the objectives of the study.

Here we will focus on the boundaries of the CWAC site, which can be downloaded from the
CWAC server, but the workflow will be exactly the same for any other polygons.

We will use the same function `uploadFeaturesToEE()` to upload our counts, but now we will
have polygons associated with these counts. The workflow is almost exactly the same as
the one we saw for pentads. Now, we will need to specify a spatial reducer in some
of the functions. This is because our pixel information will now need to be summarized to a
single value for each polygon.

The first thing we need to do is to download the polygons corresponding to the sites
we are interested in. We can do this with the function `getCwacSiteBoundary()`.

```{r eval=FALSE}

# Download our counts for the Black Duck just in case they got lost
counts <- listCwacSpp() %>% 
  filter(Common_species == "African Black",
         Common_group == "Duck") %>% 
  pull("SppRef") %>% 
  getCwacSppCounts()

# Then let's extract the boundaries of the CWAC sites in our data
# At the moment getCwacSiteBoundary() can only retrieve boundaries from sites of
# one province/country at at a time

# First identify the countries in our data
unique(counts$Country)

# It's only South Africa at the time, so we can pull all the sites at once. Otherwise,
# we would just repeat the process for the different countries.
sites <- unique(counts$LocationCode)

boundaries <- getCwacSiteBoundary(loc_code = sites,
                                  region_type = "country",
                                  region = "South Africa")

# Add boundaries to the count data
counts_bd <- counts %>% 
  dplyr::left_join(boundaries, by = "LocationCode")

```

You may notice that some of the site boundaries might not be yet available in 
the database. This will hopefully be fixed soon! For the purpose of this tutorial
we will focus on the those sites for which boundaries are available, but you probably
shouldn't do this in your own analysis!


```{r eval=FALSE}
# Filter counts to those with boundary geometry
counts_bd <- counts_bd %>% 
  dplyr::filter(!sf::st_is_empty(geometry))
```

After this we just need to follow a very similar procedure as before to annotate
with layers from the Google Earth Engine catalog.

```{r eval=FALSE}

# Select variables and format them
counts_to_upload <- counts_bd %>%
  dplyr::select(ID, StartDate) %>%
  mutate(StartDate = as.character(StartDate))

# Set an ID for your remote asset (data in GEE)
assetId <- file.path(ee_get_assethome(), 'cwac_counts_bd')

# Upload to GEE (if not done already - do this only once)
ee_counts <- uploadFeaturesToEE(feats = counts_to_upload,
                                asset_id = assetId,
                                load = TRUE)

# Annotate with surface water occurrence. Now we need to specify a reducer function to
# summarise our variable per polygon. We will use the mean in this case
counts_water <- addVarEEimage(ee_feats = ee_counts,
                              image = "JRC/GSW1_3/GlobalSurfaceWater",
                              bands = "occurrence",
                              reducer = "mean") # Note this reducer to summarize per polygon

# We could similarly annotate with an image collection. In this case, we use
# the minimum temperature from TerraClimate and we will have to specify a 
# spatial reducer, in addition to the temporal reducer specified earlier. We will
# use the min minimum temperature.
counts_tmmn <- addVarEEcollection(ee_feats = ee_counts, 
                                  collection = "IDAHO_EPSCOR/TERRACLIMATE",
                                  dates = c("2010-01-01", "2011-01-01"),
                                  temp_reducer = "mean",
                                  spt_reducer = "min",
                                  bands = "tmmn")



```

The other annotating functions will work similarly. We just need to think of using
the appropriate spatial reducer.

## INSTRUCTIONS TO CONTRIBUTE CODE

First clone the repository to your local machine:

- In RStudio, create a new project
- In the ‘Create project’ menu, select ‘Version Control’/‘Git’
- Copy the repository URL (click on the ‘Code’ green button and copy the
  link)
- Choose the appropriate directory and ‘Create project’
- Remember to pull the latest version regularly

For site owners:

There is the danger of multiple people working simultaneously on the
project code. If you make changes locally on your computer and, before
you push your changes, others push theirs, there might be conflicts.
This is because the HEAD pointer in the main branch has moved since you
started working.

To deal with these lurking issues, I would suggest opening and working
on a topic branch. This is a just a regular branch that has a short
lifespan. In steps:

- Open a branch at your local machine
- Push to the remote repo
- Make your changes in your local machine
- Commit and push to remote
- Create pull request:
  - In the GitHub repo you will now see an option that notifies of
    changes in a branch: click compare and pull request.
- Delete the branch. When you are finished, you will have to delete the
  new branch in the remote repo (GitHub) and also in your local machine.
  In your local machine you have to use Git directly, because apparently
  RStudio doesn´t do it:
  - In your local machine, change to master branch.
  - Either use the Git GUI (go to branches/delete/select branch/push).
  - Or use the console typing ‘git branch -d your_branch_name’.
  - It might also be necessary to prune remote branches with ‘git remote
    prune origin’.

Opening branches is quick and easy, so there is no harm in opening
multiple branches a day. However, it is important to merge and delete
them often to keep things tidy. Git provides functionality to deal with
conflicting branches. More about branches here:

<https://git-scm.com/book/en/v2/Git-Branching-Branches-in-a-Nutshell>

Another idea is to use the ‘issues’ tab that you find in the project
header. There, we can identify issues with the package, assign tasks and
warn other contributors that we will be working on the code.
